Do we need detailed demographic data to forecast plant population responses to climate change?

### 2015-07-10 manuscript/appA.Rmd
adds QBM full model description and STAN code

@@

\begin{align}
x_{ijQ,t+1} &= \gamma^{G}_{j,t} + \phi^{G}_{jQ} + \beta^{G}_{j,t}x_{ij,t} + \omega^{G}_{j}w_{ij,t} + \nu^{S}_{j}w_{ij,t}x_{ij,t} + \theta^{G}_{jk}C_{k,t} \\
Y^{G}_{ijQ,t} &\sim \text{Normal}(x_{ijQ,t+1}, \sigma_{j})
\end{align}

@@

**Data, process, and parameter models.**
\begin{align}
&\mathtt{data} \qquad &\textbf{y} \sim \text{Normal}(\boldsymbol{\mu}, \boldsymbol{\sigma}^2) \\
&\mathtt{process} \qquad &\mu_{iQ,t+1} = \gamma_{t} + \phi_{Q} + \beta_{t}x_{i,t} + \omega w_{i,t} + \nu w_{i,t}x_{i,t} + \theta_{k}C_{k,t} \\
& \qquad & \sigma^2_{iQ,t} = ae^{b\mu_{iQ,t+1}} \\
&\mathtt{parameters} \qquad &\gamma_{t} \sim \text{Normal}(\tilde{\gamma}, \sigma_\gamma^2)\\
& \qquad & \phi_{Q} \sim \text{Normal}(0,\sigma_Q^2) \\
& \qquad & \beta_{t} \sim \text{Normal}(\tilde{\beta}, \sigma_\beta^2) \\
& \qquad & \omega \sim \text{Normal}(0, 0.001) \\
& \qquad & \nu \sim \text{Normal}(0, 0.001) \\
& \qquad & \boldsymbol{\theta} \sim \text{Normal}(0, 0.001) \\
& \qquad & \tilde{\gamma} \sim \text{Normal}(0, 0.001) \\
& \qquad & \tilde{\beta} \sim \text{Normal}(0, 0.001) \\
& \qquad & a \sim \text{Normal}(0, 1000) \\
& \qquad & b \sim \text{Normal}(0, 1000) \\
& \qquad & \sigma_Q^2 \sim \text{Uniform}(0, 1000) \\
& \qquad & \sigma_\gamma^2 \sim \text{Uniform}(0, 1000) \\
& \qquad & \sigma_\beta^2 \sim \text{Uniform}(0, 1000)
\end{align}

**Full expression of posterior and joint distributions**
\begin{align}
[\boldsymbol{\gamma}, \tilde{\gamma}, \boldsymbol{\phi}, \boldsymbol{\beta}, \tilde{\beta}, \omega, \boldsymbol{\theta}, \sigma^2, \sigma_Q^2, \sigma_\beta^2] &\propto \\
&\prod_{t=1}^T \prod_{i=1}^n [y_{iQ,t} | \gamma_{t}, \phi_{Q}, \beta_{t}, \omega, \boldsymbol{\theta}, \sigma^2] [\gamma_{t} | \tilde{\gamma}] [\beta_{t} | \tilde{\beta}, \sigma_\beta^2] \times \\
&\prod_{Q=1}^{Q_{tot}} [\phi_{Q} | \sigma_Q^2] \times \\
&[\boldsymbol{\theta}] [\omega] [\tilde{\gamma}] [\tilde{\beta}] [\sigma_Q^2] [\sigma_\beta^2]
\end{align}

@@

**Data, process, and parameter models**
\begin{align}
&\mathtt{data} \qquad &\textbf{y} \sim \text{NegBin}(\boldsymbol{\lambda},\zeta) \\
&\mathtt{process} \qquad &\lambda_{jq,t+1} = A'_{q,t}e^{(\gamma_{t} + \phi_{Q} + \theta_{k}C_{k,t} + \omega\sqrt{A'_{q,t}})} \\
&\mathtt{parameters} \qquad &\gamma_{t} \sim \text{Normal}(\tilde{\gamma}, \sigma_\gamma^2)\\
& \qquad & \phi_{Q} \sim \text{Normal}(0,\sigma_Q^2) \\
& \qquad & \omega \sim \text{Uniform}(-100, 100) \\
& \qquad & \boldsymbol{\theta} \sim \text{Uniform}(-10, 10) \\
& \qquad & \tilde{\gamma} \sim \text{Uniform}(-300, 300) \\
& \qquad & \sigma_Q^2 \sim \text{Cauchy}(0, 5) \\
& \qquad & \sigma_\gamma^2 \sim \text{Cauchy}(0, 5)
\end{align}

@@

>>>>

Parameter to model notation keys
--------------------------------

Table:  Definition of parameters and model notation for IPM survival and growth models.

| Parameter | Definition | Notation in statistical results tables |
| --------- | ---------- | -------------------------------------- |
| $\gamma_{t}$ | intercept for year *t* | ``a[t]`` where ``t`` is the numeric year |
| $\tilde{\gamma}$ | global intercept | ``a_mu`` |
| $\phi_{Q}$ | random effect of quadrat group *Q* | ``gint[Q]`` where ``Q`` is the quadrat group |
| $\beta_{t}$ | size effect for year *t* | ``b1[t]`` where ``t`` is the numeric year |
| $\tilde{\beta}$ | global size effect | ``b1_mu`` |
| $\gamma$ | effect of crowding | ``w[1]`` |
| $\nu$ | crowding $\times$ size effect | ``w[2]`` |
| $\theta_{k}$ | effect of climate covariate *k* | ``b2[k]`` where ``k`` is the climate effect |
| *a* | first parameter for growth variance | ``tau`` |
| *b* | second parameter for growth variance | ``tauSize`` |

Table:  Definition of parameters and model notation for IPM recruitment model.

| Parameter | Definition | Notation in statistical results tables |
| --------- | ---------- | -------------------------------------- |
| $\gamma_{t}$ | intercept for year *t* | ``a[t]`` where ``t`` is the numeric year |
| $\tilde{\gamma}$ | global intercept | ``a_mu`` |
| $\phi_{Q}$ | random effect of quadrat group *Q* | ``gint[Q]`` where ``Q`` is the quadrat group |
| $\gamma$ | effect of plot cover | ``dd`` |
| $\theta_{k}$ | effect of climate covariate *k* | ``b2[k]`` where ``k`` is the climate effect |
| *p* | mixing fraction for effective plot cover | ``u`` |
| $\zeta$ | size parameter for negative binimial likelihood | ``theta`` |

Table:  Definition of parameters and model notation for quadrat based model.

| Parameter | Definition | Notation in statistical results tables |
| --------- | ---------- | -------------------------------------- |
| $\gamma_{t}$ | intercept for year *t* | ``a[t]`` where ``t`` is the numeric year |
| $\tilde{\gamma}$ | global intercept | ``a_mu`` |
| $\phi_{Q}$ | random effect of quadrat group *Q* | ``gint[Q]`` where ``Q`` is the quadrat group |
| $\beta_{t}$ | cover effect for year *t* | ``b1[t]`` where ``t`` is the numeric year |
| $\tilde{\beta}$ | global cover effect | ``b1_mu`` |
| $\theta_{k}$ | effect of climate covariate *k* | ``b2[k]`` where ``k`` is the climate effect |
| $\tau$ | model variance in log normal likelihood | ``tau`` |

Table: Climate effect key.

| Integer ID (*k*) | Climate covariate |
| ---------------- | ----------------- |
| 1 | pptLag |
| 2 | ppt1 |
| 3 | ppt2 |
| 4 | TmeanSpr1 |
| 5 | TmeanSpr2 |
| 6 | ppt1 $\times$ TmeanSpr1 |
| 7 | ppt2 $\times$ TmeanSpr2 |
| 8 | pptLag $\times$ size |
| 9 | ppt1 $\times$ size |
| 10 | ppt2 $\times$ size |
| 11 | TmeanSpr1 $\times$ size |
| 12 | TmeanSpr2 $\times$ size |

@@

\begin{align}
x_{ijQ,t+1} &= \gamma^{G}_{j,t} + \phi^{G}_{jQ} + \beta^{G}_{j,t}x_{ij,t} + \omega^{G}_{j}w_{ij,t} + \nu^{G}_{j}w_{ij,t}x_{ij,t} + \theta^{G}_{jk}C_{k,t} \\
Y^{G}_{ijQ,t} &\sim \text{Normal}(x_{ijQ,t+1}, \sigma_{j})
\end{align}

@@

**Data, process, and parameter models.**
\begin{align}
&\mathtt{data} \qquad &\textbf{y} \sim \text{Normal}(\boldsymbol{\mu}, \boldsymbol{\sigma}^2) \\
&\mathtt{process} \qquad &\mu_{iQ,t+1} = \gamma_{t} + \phi_{Q} + \beta_{t}x_{i,t} + \omega w_{i,t} + \nu w_{i,t}x_{i,t} + \theta_{k}C_{k,t} \\
& \qquad & \sigma^2_{iQ,t} = ae^{b\mu_{iQ,t+1}} \\
&\mathtt{parameters} \qquad &\gamma_{t} \sim \text{Normal}(\tilde{\gamma}, \sigma_\gamma^2)\\
& \qquad & \phi_{Q} \sim \text{Normal}(0,\sigma_Q^2) \\
& \qquad & \beta_{t} \sim \text{Normal}(\tilde{\beta}, \sigma_\beta^2) \\
& \qquad & \omega \sim \text{Normal}(0, 0.001) \\
& \qquad & \nu \sim \text{Normal}(0, 0.001) \\
& \qquad & \boldsymbol{\theta} \sim \text{Normal}(0, 0.001) \\
& \qquad & \tilde{\gamma} \sim \text{Normal}(0, 0.001) \\
& \qquad & \tilde{\beta} \sim \text{Normal}(0, 0.001) \\
& \qquad & a \sim \text{Normal}(0, 0.001) \\
& \qquad & b \sim \text{Normal}(0, 0.001) \\
& \qquad & \sigma_Q^2 \sim \text{Uniform}(0, 1000) \\
& \qquad & \sigma_\gamma^2 \sim \text{Uniform}(0, 1000) \\
& \qquad & \sigma_\beta^2 \sim \text{Uniform}(0, 1000)
\end{align}

**Full expression of posterior and joint distributions**
\begin{align}
[\boldsymbol{\gamma}, \tilde{\gamma}, \boldsymbol{\phi}, \boldsymbol{\beta}, \tilde{\beta}, \omega, \nu, \boldsymbol{\theta}, a, b, \sigma_Q^2, \sigma_\beta^2] &\propto \\
&\prod_{t=1}^T \prod_{i=1}^n [y_{iQ,t} | \gamma_{t}, \phi_{Q}, \beta_{t}, \omega, \nu, \boldsymbol{\theta}, a, b] [\gamma_{t} | \tilde{\gamma}] [\beta_{t} | \tilde{\beta}, \sigma_\beta^2] \times \\
&\prod_{Q=1}^{Q_{tot}} [\phi_{Q} | \sigma_Q^2] \times \\
&[\boldsymbol{\theta}] [\omega] [\nu] [\tilde{\gamma}] [\tilde{\beta}] [\sigma_Q^2] [\sigma_\beta^2] [a] [b]
\end{align}

@@

**Data, process, and parameter models**
\begin{align}
&\mathtt{data} \qquad &\textbf{y} \sim \text{NegBin}(\boldsymbol{\lambda},\zeta) \\
&\mathtt{process} \qquad &\lambda_{jq,t+1} = A'_{q,t}e^{(\gamma_{t} + \phi_{Q} + \theta_{k}C_{k,t} + \omega\sqrt{A'_{q,t}})} \\
&\mathtt{parameters} \qquad &\gamma_{t} \sim \text{Normal}(\tilde{\gamma}, \sigma_\gamma^2)\\
& \qquad & \phi_{Q} \sim \text{Normal}(0,\sigma_Q^2) \\
& \qquad & \omega \sim \text{Uniform}(-100, 100) \\
& \qquad & \boldsymbol{\theta} \sim \text{Uniform}(-10, 10) \\
& \qquad & \tilde{\gamma} \sim \text{Uniform}(-300, 300) \\
& \qquad & \sigma_Q^2 \sim \text{Cauchy}(0, 5) \\
& \qquad & \sigma_\gamma^2 \sim \text{Cauchy}(0, 5) \\
& \qquad & \zeta \sim \text{Uniform}(0, 100) \\
& \qquad & u \sim \text{Uniform}(0, 1)
\end{align}

**Full expression of posterior and joint distributions**
\begin{align}
[\boldsymbol{\gamma}, \tilde{\gamma}, \boldsymbol{\phi}, \omega, \boldsymbol{\theta}, \sigma_Q^2, \zeta, u] &\propto \\
&\prod_{t=1}^T \prod_{i=1}^n [y_{iQ,t} | \gamma_{t}, \phi_{Q}, \omega, \boldsymbol{\theta}, \zeta, u] [\gamma_{t} | \tilde{\gamma}] \times \\
&\prod_{Q=1}^{Q_{tot}} [\phi_{Q} | \sigma_Q^2] [\boldsymbol{\theta}] [\omega] [\tilde{\gamma}] [\sigma_Q^2] [\zeta] [u]
\end{align}

@@

Quadrat based model (QBM)
-------------------------
The model for quadrat cover ($\mu$) change from time $t$ to $t+1$ is

\begin{align}
\mu_{jq,t+1} &= \gamma_{j,t} + \phi_{jQ} + \beta_{j,t}x_{jq,t} + \theta_{jk}C_{k,t} \\
Y_{jq,t+1} &\sim \text{LogNormal}(\mu_{jq,t+1}, \tau{j}) \text{T}[0,1]
\end{align}

where $x_{jq,t}$ is the log of species' $j$ proportional cover in quadrat $q$ at time $t$ and all other parameters are as in the individual-level growth model (Eq. 3). Again, note that the climate covariate vector (**C**) includes the climate$\times$cover interaction. The log normal likelihood includes a truncation (T[0,1]) to ensure that predicted values do not exceed 100% cover.

**Data, process, and parameter models.**
\begin{align}
&\mathtt{data} \qquad &\textbf{y} \sim \text{LogNormal}(\boldsymbol{\mu}, \boldsymbol{\tau}) \text{T}[0,1] \\
&\mathtt{process} \qquad &\mu_{iQ,t+1} = \gamma_{t} + \phi_{Q} + \beta_{t}x_{i,t} + \theta_{k}C_{k,t} \\
&\mathtt{parameters} \qquad &\gamma_{t} \sim \text{Normal}(\tilde{\gamma}, \sigma_\gamma^2)\\
& \qquad & \phi_{Q} \sim \text{Normal}(0,\sigma_Q^2) \\
& \qquad & \beta_{t} \sim \text{Normal}(\tilde{\beta}, \sigma_\beta^2) \\
& \qquad & \boldsymbol{\theta} \sim \text{Uniform}(-10, 10) \\
& \qquad & \tilde{\gamma} \sim \text{Uniform}(-300, 300) \\
& \qquad & \tilde{\beta} \sim \text{Uniform}(-100, 100) \\
& \qquad & \sigma_Q^2 \sim \text{Cauchy}(0, 5) \\
& \qquad & \sigma_\gamma^2 \sim \text{Cauchy}(0, 5) \\
& \qquad & \sigma_\beta^2 \sim \text{Cauchy}(0, 5) \\
& \qquad & \tau \sim \text{Cauchy}(0, 5)
\end{align}

**Full expression of posterior and joint distributions**
\begin{align}
[\boldsymbol{\gamma}, \tilde{\gamma}, \boldsymbol{\phi}, \boldsymbol{\beta}, \tilde{\beta}, \boldsymbol{\theta}, \sigma_Q^2, \sigma_\beta^2, \tau] &\propto \\
&\prod_{t=1}^T \prod_{i=1}^n [y_{iQ,t} | \gamma_{t}, \phi_{Q}, \beta_{t}, \boldsymbol{\theta}, \tau] [\gamma_{t} | \tilde{\gamma}] [\beta_{t} | \tilde{\beta}, \sigma_\beta^2] \times \\
&\prod_{Q=1}^{Q_{tot}} [\phi_{Q} | \sigma_Q^2] \times \\
&[\boldsymbol{\theta}] [\omega] [\tilde{\gamma}] [\tilde{\beta}] [\sigma_Q^2] [\sigma_\beta^2] [\tau]
\end{align}

**Stan code for model**
```{r qbm_stan, echo=TRUE, eval=FALSE}
data{
  int<lower=0> N; // observations
  int<lower=0> Yrs; // years
  int<lower=0> yid[N]; // year id
  int<lower=0> Covs; // climate covariates
  int<lower=0> G; // groups
  int<lower=0> gid[N]; // group id
  real<lower=0,upper=1> Y[N]; // observation vector
  matrix[N,Covs] C; // climate matrix
  vector[N] X; // size vector
}
parameters{
  real a_mu;
  vector[Yrs] a;
  real b1_mu;
  vector[Yrs] b1;
  vector[Covs] b2;
  vector[G] gint;
  real<lower=0> sig_a;
  real<lower=0> sig_b1;
  real<lower=0> sig_G;
  real<lower=0> tau;
}
transformed parameters{
  real mu[N];
  vector[N] climEff;
  climEff <- C*b2;
  for(n in 1:N)
    mu[n] <- a[yid[n]] + gint[gid[n]] + b1[yid[n]]*X[n] + climEff[n];
}
model{
  // Priors
  a_mu ~ uniform(-300,300);
  b1_mu ~ uniform(-100,100);
  sig_a ~ cauchy(0,5);
  sig_b1 ~ cauchy(0,5);
  sig_G ~ cauchy(0,5);
  gint ~ normal(0, sig_G);
  b2 ~ uniform(-10,10);
  a ~ normal(a_mu, sig_a);
  b1 ~ normal(b1_mu, sig_b1);
  tau ~ cauchy(0,5);

  //Likelihood
  Y ~ lognormal(mu, tau);
}
```
